The curious world of constructive mathematics

### 2022-09-08 algar-lecture-notes.tex
Abstract for first lecture on toposes

>>>>

\begin{intro}
Just as the easiest and shortest path between two truths of the real domain
often passes through the complex domain, sometimes the easiest and most
conceptual path to a result observes that the claim can also be formulated in
the internal language of an alternate mathematical universe -- an alternate
topos -- where the proof can be simple, direct and make use of the finer
distinctions provided by intuitionistic logic.

As such, toposes are the workhorse of many modern applications of constructive
mathematics to mathematics in general. Toposes originated in algebraic
geometry, where they have been used as generalized spaces (allowing for a
nontrivial category of opens instead of only a partially ordered set of opens),
before category theorists observed that there is also a strong logical aspect
to toposes: Toposes can be pictured as mathematical universes -- universes in
which we can do mathematics much in the same way as in the so-called standard
topos of sets and maps, the single topos in which most mathematicians spend all
their professional life in.

Besides the standard topos, in which mathematics unfolds exactly as known and
classical logic reigns, there is a colorful host of alternate toposes -- some
intimately related and useful to the mathematics of the standard topos, some
less so and important to computability theorists and logicians. In most
alternate toposes, only intuitionistic logic is sound and it is a plain fact of
the matter that the law of excluded middle and the axiom of choice fail.

In the lecture we will focus on precise definitions, examples of toposes, and
logical applications to mathematics in general, laying the groundwork for the
fourth lecture on generic models and carefully elucidating why it is that most
toposes validate only intuitionistic logic. References include
\cite{blechschmidt:filmat,leinster:introduction,streicher:ctcl,moerdijk-maclane:sheaves-logic,goldblatt:topoi,johnstone:elephant}
(and \cite{blechschmidt:phd} for applications in algebraic geometry). These
themes constitute just a tiny part of topos theory. Exposition on Olivia
Caramello's grand topos-theoretic bridge-building program~\cite{caramello:tst},
on applications of toposes to the meta-analysis of
logic~\cite{lambek-scott:hocl} and how topos theory can be used to reconcile
seemingly disparate positions in the philosophy of
mathematics~\cite{lambek:incompatible,couture-lambek:reflections} will take
place in more informal settings than the lecture.
\end{intro}
